---
title: Linear regression and sample size
author: Peter Ralph
date: "`r date()`"
---

```{r setup, include=FALSE}
fig.dim <- 5
library(knitr)
library(colorspace) # for adjustcolor
opts_chunk$set(
    fig.height=fig.dim,
    fig.width=fig.dim,
    fig.align='center'
)
```

We'd like to know how accurately we can estimate
the slope of a linear relationship
as a function of sample size.

Oh, and we're stuck on a desert island with no internet access
or statistics books
and we forgot all our math.

So, let's find out by simulation!


One data set
------------

Let's simulate some samples,
fit a linear model,
and make a plot
just to check everything looks right.


Simulated data:

```{r sim_first}
n <- 10
slope <- 0.2
noise.sd <- 0.25
xy <- data.frame( x=rnorm(n))
xy$y <- xy$x * slope + noise.sd*rnorm(n)
summary(xy)
```

Now, fit a linear model:

```{r fit_model}
xy.lm <- lm( y ~ x, data=xy )
summary(xy.lm)
```

Look at the results:

```{r first_plot}
plot( y ~ x, data=xy )
abline(coef(xy.lm),col='red')
```


Many data sets:
---------------

We'd like to do the above many times for a given sample size.


```{r sim_many, fig.width=2*fig.dim}
n <- 10
slope <- 0.2
noise.sd <- 0.25
nreps <- 1000
many.lms <- replicate( nreps, {
        xy <- data.frame( x=rnorm(n))
        xy$y <- xy$x * slope + noise.sd*rnorm(n)
        xy.lm <- lm( y ~ x, data=xy )
        return( coef(xy.lm) )
    } )
# hist the intercept and slope
layout(t(1:2))
hist( many.lms[1,], main="intercept" )
hist( many.lms[2,], main="slope" )
```

Let's just look at a bunch of those lines, for fun:

```{r many_lines}
plot(0,type='n',xlab='',ylab='',xlim=c(-1,1),ylim=c(-1,1))
for (k in 1:100) {
    abline( many.lms[1,k], many.lms[2,k], col=adjustcolor("black",0.1) )
}
```


Systematic comparison
---------------------

First thing: *make a function*.

```{r setup_sims}
sim.many.lms <- function (n, nreps=1000) {
    many.lms <- replicate( nreps, {
            xy <- data.frame( x=rnorm(n))
            xy$y <- xy$x * slope + noise.sd*rnorm(n)
            xy.lm <- lm( y ~ x, data=xy )
            return( coef(xy.lm) )
        } )
}
```

Now, apply it to the sample sizes of interest:

```{r run_sims, cache=TRUE}
n.vals <- floor(seq(10,500,length.out=10))
sim.lms <- lapply( n.vals, sim.many.lms )
```
```{r plot_sims, fig.width=2*fig.dim}
layout(t(1:2))
for (k in seq_along(n.vals)) {
    hist(sim.lms[[k]][1,], main=paste("intercepts, n=",n.vals[k]))
    hist(sim.lms[[k]][2,], main=paste("slopes, n=",n.vals[k]))
}
```

OK, let's summarize the results on slope:
```{r show_slope, fig.width=2*fig.dim}
layout(t(1:2))
boxplot( lapply( sim.lms, function (x) x[2,] ), names=n.vals )
plot(n.vals, sapply( sim.lms, function (x) sd(x[2,]) ), ylab="SD of slope" )
```
