----
Title: "Generating sequences without repeats"
Author: "Peter Ralph"
Date: "`r date()`"
----

**Goal:**
To write a function that takes a single argument, $n$,
and returns all strings of length $n$
composed of A, C, G, and T
in which no two adjacent letters are the same.

To start, and for testing purposes, let's compile the list by hand for small $n$.
For instance, `truth[[2]]` will have the right answer for `n=2`:
```{r right_answer}
truth <- list(
              c("A", "C", "G", "T"),
              c("AC", "AG", "AT",
                "CA", "CG", "CT",
                "GA", "GC", "GT",
                "TA", "TC", "TG"),
              c("ACA", "ACG", "ACT",
                "AGA", "AGC", "AGT",
                "ATA", "ATC", "ATG",
                "CAC", "CAG", "CAT",
                "CGA", "CGC", "CGT",
                "CTA", "CTC", "CTG",
                "GAC", "GAG", "GAT",
                "GCA", "GCG", "GCT",
                "GTA", "GTC", "GTG",
                "TAC", "TAG", "TAT",
                "TCA", "TCG", "TCT",
                "TGA", "TGC", "TGT"))
```

As a check, the correct number of these should be `4 * 3^{(n-1)}` --
four for the first letter, and three for each subsequent one.
```{r check_truth_lengths}
4 * (3^(0:2))
sapply(truth, length)
```

The procedure of making those lists suggests an elegant idea:
construct the length $(n+1)$ strings by prepending a letter to the length $n$ strings,
but omitting the ones that would make a repeat.
```{r the_function}
sequences <- function (n) {
    bases <- c("A", "C", "G", "T")
    if (n==1) {
        out <- bases
    } else {
        out <- c()
        smaller <- sequences(n-1)
        for (x in bases) {
            omit <- (substr(smaller, 1, 1) == x)
            out <- c(out, paste0(x, smaller[!omit]))
        }
    }
    return(out)
}
```

Now let's check this:
```{r check_function}
all(sequences(1) == truth[[1]])
all(sequences(2) == truth[[2]])
all(sequences(3) == truth[[3]])
results <- sapply(1:8, sequences)
plot(sapply(results, length), xlab='n', ylab='length of result')
lines(1:8, 4*3^(0:7), col='red')
```


